---
title: "Scripts for the Data Paper"
author: "Florencia Grattarola"
date: "12 June 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(extrafont)
library(raster)
```

This post has all the scripts used to generate the figures of the Data Paper **“Biodiversidata: An Open-Access Biodiversity Database for Uruguay”**, published in the Biodiversity Data Journal. The dataset **Biodiversidata_1.0.0** provides primary biodiversity data on extant Amphibia, Reptilia, Aves and Mammalia species recorded within the country area of Uruguay. This is the largest and most geographically and taxonomically comprehensive database of Uruguayan tetrapod species available to date, and it represents the first open repository of the country.  

You can find the data here  [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.2650169.svg)](https://doi.org/10.5281/zenodo.2650169)

---

## Enough talking! Let's explore the data!

To run this code you will need the following R packages:

```r 
library(tidyverse)
library(extrafont)
library(dismo)
```  
  
We will see the data's:  
1. [Taxonomic coverage](#taxonomic-coverage)  
2. [Temporal coverage](#temporal-coverage)  
3. [Geographic coverage](#geographic-coverage)  

Let's start by getting the database. We could read the file like this: 
```r 
Biodiversidata <- read_delim('Biodiversidata_1.0.0.tsv', delim = '\t')
```  

But we are going to include all the column names and types to avoid having any reading errors on our file.

```{r echo=TRUE}
Biodiversidata <- read_delim('Biodiversidata_1.0.0.tsv', 
                             col_types = cols(occurrenceID = col_character(),
                                              scientificName = col_character(),
                                              scientificNameAuthorship = col_character(),
                                              vernacularName = col_character(),
                                              kingdom = col_character(),
                                              phylum = col_character(),
                                              class = col_character(),
                                              order = col_character(),
                                              family = col_character(),
                                              genus = col_character(),
                                              specificEpithet = col_character(),
                                              infraspecificEpithet = col_character(),
                                              countryCode = col_character(),
                                              stateProvince = col_character(),
                                              verbatimLocality = col_character(),
                                              decimalLatitude = col_number(),
                                              decimalLongitude = col_number(),
                                              georeferenceSources = col_character(),
                                              georeferencedBy = col_character(),
                                              eventDate = col_date(format = ""),
                                              year = col_number(),
                                              month = col_number(),
                                              day = col_number(),
                                              basisOfRecord = col_character(),
                                              institutionCode = col_character(),
                                              collectionCode = col_character(),
                                              catalogNumber = col_character(),
                                              recordedBy = col_character(),
                                              recordNumber = col_character(),
                                              identifiedBy = col_character(),
                                              dynamicProperties = col_character(),
                                              associatedReferences  = col_character()), delim = '\t')
```


---


Now that we are ready! Let's explore the data.  

### What's the total amount of records in the database?

```{r}
Biodiversidata %>% nrow()
```

### What's the total amount of species in the database?

```{r}
Biodiversidata %>% distinct(scientificName) %>% nrow()
```


## 1) Taxonomic coverage

The database includes records representing 129 families, 446 genera and 673 species: 51 amphibians, 68 reptiles, 437 birds, and 117 mammals. 

### What's the number of records per class?

```{r}
Biodiversidata %>% 
  group_by(class) %>% 
  count()  
```

### Let's plot this

```{r}
Biodiversidata %>% 
  group_by(class) %>% 
  count() %>% 
  ggplot(aes(x='', y=n, fill=class)) +
  geom_bar(width = 0.5, stat = "identity") +
  labs(x='', y= 'Number of Species', fill = '') +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 70000,10000)) +
  scale_fill_brewer(palette ='Spectral')+
  theme(text=element_text(family='Calibri', size = 12)) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)))
```

### What's the number of species per class?

```{r}
Biodiversidata %>% 
  distinct(scientificName, class) %>% 
  group_by(class) %>% 
  count() 
```

### and the plot?

```{r}
Biodiversidata %>% 
  distinct(scientificName, class) %>% 
  group_by(class) %>% 
  count() %>% 
  ggplot(aes(x='', y=n, fill=class)) +
  geom_bar(width = 0.5, stat = "identity") +
  labs(x='', y= 'Number of Species', fill = '') +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 1000, 100)) +
  scale_fill_brewer(palette ='Spectral')+
  theme(text=element_text(family='Calibri', size = 12)) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)))
```

The taxonomic coverage is uneven!

## 2) Temporal coverage

The records included in Biodiversidata cover samples reported in Uruguay during the period of 1806–2018. How are these records distributed over time within the groups?

```{r message = FALSE}
ggplot(Biodiversidata, aes(x=year, y=order, color=class)) +
  geom_point(show.legend = FALSE) +
  scale_colour_brewer(palette ='Spectral')+
  facet_grid(class~., scales = "free", space= 'free_y', switch='x' ,drop=TRUE) +
  theme_bw() +
  theme(text=element_text(family='Calibri', size=12)) +
  theme(strip.text.y = element_text(size=10, angle=0)) +
  labs(x='Year', y='Order', color = '') 
```


## 3) Geographic coverage

To finish let's map the records. 
First, we are going to download Uruguay's shapefile data from [GADM](http://gadm.org/) using the ISO country code 'UY' and the level 1 for administrative boundaries. 

```{r}
Uruguay <- getData("GADM", country="UY", level=1)
```


Now we map the occurrence records separated by class.

```{r message=FALSE}
ggplot() + 
  geom_polygon(data=Uruguay, aes(long,lat,group=group), fill="white") +
  geom_path(data=Uruguay, aes(long,lat, group=group), color="grey", size=0.1) +
  geom_point(data = Biodiversidata, aes(x=decimalLongitude, y = decimalLatitude, color = factor(class))) +
  scale_color_brewer(palette ='Spectral') +
  labs(x='', y= '', color = '') +
  theme_bw() +
  theme(text=element_text(family='Calibri', size = 12))
```


## And that's all !

---
  

**If you use the data please cite as:**  
Florencia Grattarola, Germán Botto, Inés da Rosa, Noelia Gobel, Enrique M. González, Javier González, Daniel Hernández, Gabriel Laufer, Raúl Maneyro, Juan A. Martínez-Lanfranco, Daniel E. Naya, Ana L. Rodales, Lucía Ziegler & Daniel Pincheira-Donoso. (2019). 'Biodiversidata: An Open-Access Biodiversity Database for Uruguay'. Biodiversity Data Journal. Available online: https://doi.org/10.5281/zenodo.2650169  
  

